---
title: "Harvest plot tutorial"
author: "Christina Xiao"
date: "2023-04-05"
output:
  prettydoc::html_pretty:
    theme: cayman
---

***
<br><br>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Rich has made an edit.

This RMarkdown document demonstrates how to create a harvest plot when there are multiple outcomes and exposures of interest within a systematic review. This helps to automate the plotting while allowing the user to change which dimensions are represented by different study characteristics.
<br><br>
Harvest plots have been defined by [Ogilvie et al., 2009](https://bmcmedresmethodol.biomedcentral.com/articles/10.1186/1471-2288-8-8) as:
<br>

>A novel and useful method for synthesising evidence about the differential effects of population-level interventions. It contributes to the challenge of making best use of all available evidence by incorporating all relevant data. The visual display assists both the process of synthesis and the assimilation of the findings. The method is suitable for adaptation to a variety of questions in evidence synthesis and may be particularly useful for systematic reviews addressing the broader type of research question which may be most relevant to policymakers.

<br>

The following tutorial is based on the harvest plot from the following publication: "Shifting towards healthier transport: Carrots or sticks? Systematic review and meta-analysis of population-level interventions" [Xiao et al., 2022](https://www.thelancet.com/journals/lanplh/article/PIIS2542-5196(22)00220-0/fulltext)

This study aimed to compare the effectiveness of carrot, stick, and combined carrot-and-stick interventions in changing different transport behaviors. We will use this data to design a harvest plot for the driving and public transport outcomes.

With this in mind, let's get plotting!
<<<<<<< HEAD

=======
>>>>>>> 9860e83984eb4fcd724ed4de41608582579f2237

<br><br>

***

# Setup
Loading R packages
```{r message=FALSE, echo=FALSE}
library(readxl)
library(tidyverse)
library(knitr)
library(ggeffects)
library(ggpubr)
library(gtsummary)
library(flextable)
library(janitor)
```


```{r message=FALSE, warning=FALSE}
library(readxl)
library(tidyverse)
library(knitr)
library(ggpubr)
```
<br><br>

***

# Load data
```{r message=FALSE, warning=FALSE}
data <- read_excel("~/Library/CloudStorage/OneDrive-UniversityofCambridge/Carrots and sticks/Harvest plot/harvest_plot_tutorial_csreview.xlsx")
```
<br><br>

***

## Defining variables 

The following defines each variable in the dataset, the first 6 rows of which are displayed using the head function.

- **study_id**: Unique study identifier <br>
- **position**: Placement of the individual study outcome in the harvest plot <br>
- **outcome_mode**: Outcome of interest in your review. This can be a single outcome or many. Here, we are interested in looking at driving and public transport <br>
- **cs_category**: Exposure or category of studies to be compared, which will be placed in separate rows<br>
- **change**: Whether there was an increase, decrease, or no change in the outcome. <br>
- **stat_sig**: Whether the difference was found to be statistically significant. This will be represented by the shading of the bars <br>
- **quality**: Indicates the study quality. We have three categories here: low, moderate, and high. This will be represented by the height of the bars<br>
- **n_interventions**: The number of intervention components, as represented by a number above each bar<br>
<br><br>

***

### Inspect data
```{r}
knitr::kable(head(data))
```
<br><br>

***

# Making changes to our variables

## We need to change some of our variables to factors:
```{r}
names <- c('quality' ,'outcome_mode', 'change','stat_sig','cs_category')
data[,names] <- lapply(data[,names] , factor)
str(data)
```

### Reordering variables
One of the changes we need to make is to reorder the change variable. Here, we're creating a new variable called **change_order**, which denotes a variable with the factored change variable ordered in the way we want it to appear in the harvest plot (Decrease, No change, Increase):
```{r}
data$change_order <- factor(data$change, levels = c("Decrease", "No change", "Increase"))
data$stat_sig <- factor(data$stat_sig, levels = c("Yes", "No", "N/A"))

```

### Creating a new variable converting study quality to a number representing the height of each bar
We also need to create a new variable representing the heights of each of our bars. Here, because I want to vary the heights based on study quality, I will need to create a new variable where the tallest height represents the high quality studies, middle represents the moderate quality studies, and the shortest represents the low quality studies:
```{r}
data <- data %>%
  mutate(quality_value = case_when(
    quality == "high"   ~ 3,
    quality == "moderate" ~ 2,
    quality == "low"    ~ 1
  ))
```
<br><br>

***

## Creating separate plots for each outcome

First, we will plot the harvest plot for driving outcomes, which we will later combine with the public transport outcomes in a separate figure. To do so, we create two new datasets called 'drive' and 'public_transport'.

```{r}
drive <- data[data$outcome_mode == 'Driving', ]
public_transport <- data[data$outcome_mode == 'Public transport', ]
```

Next, we do the same for the intervention categories of interest that we want to compare, which will represent the different rows in the figure. In this example, these are carrot, carrot and stick, and stick interventions:
```{r}
carrot <-data[data$cs_category == 'Carrot', ]
carrotandstick <-data[data$cs_category == 'Carrot and stick', ]
stick <-data[data$cs_category == 'Stick', ]
```
<br><br>

***

## Setting the position of bars representing each study
We need to also set the position of each of the bars. To do so, let's first assess how many studies are in each row and column:
```{r}
kable(table(drive$cs_category,drive$change_order))
kable(table(public_transport$cs_category,public_transport$change_order))
```
<br>
We can see here that the maximum number of studies is 20, as found in the table with Carrot interventions that were associated with a decrease in Driving outcomes.This means the study furthest to the right of the plot should be labelled with position 20, and those to the left of the plot will be labelled with position 1. 

There are different ways to order the positions, for instance, by the height of each bar and then shading/colour. If you choose to add labels above each bar, e.g., representing the number of intervention components, then the bars can be ordered this way too. If you have not done so, create a new variable called **position** with your desired order.

<br><br>

***

## Driving harvest plot
The following code plots the harvest plot for the driving outcome.

To set the colour scheme:
```{r}
cbPalette <- c("#DCDCDC", "#A9A9A9", "#696969")
```

We start building the harvest plot, specifying that the colour of each bar is represented by the variable **stat_significant**, the height of each bar by **quality_value**, and the position of each bar by **position**. We then add bars to represent each study.

```{r}
drive_a<- ggplot(drive, aes(fill=stat_sig, y=quality_value, x=Position, label = study_id)) + 
  geom_bar(position="dodge", stat="identity") + scale_fill_manual(values=cbPalette, na.translate=FALSE) 
``` 

Next, we use the facetgrid function to specify that we want **cs_category** as our rows and **change_order** as our columns.
```{r}
drive_b<- drive_a + facet_grid(cs_category~change_order,  
   switch = "y", space='free_x') + theme_minimal(base_size = 12) 
```   

The following code formats the figure so it's easier to distinguish between the different rows and columns:
```{r} 
drive_c <- drive_b +  theme(strip.placement = "outside")+ scale_y_continuous("", limits=c(0,3.5)) +
  theme(plot.title = element_text(size = 14, face = "bold"),
        legend.title=element_text(size=12,  face = "bold"), 
        legend.text=element_text(size=10)) +
  annotate("segment", x=-Inf, xend=Inf, y=-Inf, yend=-Inf)+
  annotate("segment", x=-Inf, xend=-Inf, y=-Inf, yend=Inf) +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank())
```

Specifying the title and legends:
```{r} 
drive_d <- drive_c + guides(fill = guide_legend("Significant", reverse = TRUE, nrow = 3)) +
  theme(legend.position="right") + labs(title="Driving")
```

Add the number of intervention components (**n_interventions**) above each bar:
```{r} 
drive_final <- drive_d +  geom_text(aes(label=n_interventions), position=position_dodge(width=0.9), vjust=-0.4)

```

The final plot should look like this:
```{r, fig.width=10,fig.height=6, warning=FALSE}
drive_final
```
<br><br>

***
## Public transport harvest plot
The following code plots the harvest plot for the public transport outcome.

We start building the harvest plot, specifying that the colour of each bar is represented by the variable **stat_significant**, the height of each bar by **quality_value**, and the position of each bar by **position**. We then add bars to represent each study.

```{r}
pt_a<- ggplot(public_transport, aes(fill=stat_sig, y=quality_value, x=position, label = study_id)) + 
  geom_bar(position="dodge", stat="identity") + scale_fill_manual(values=cbPalette, na.translate=FALSE) 
``` 

Next, we use the facetgrid function to specify that we want **cs_category** as our rows and **change_order** as our columns.
```{r}
pt_b<- pt_a + facet_grid(cs_category~change_order,  
   switch = "y", space='free_x') + theme_minimal(base_size = 12) 
```   

The following code formats the figure so it's easier to distinguish between the different rows and columns:
```{r} 
pt_c <- pt_b +  theme(strip.placement = "outside")+ scale_y_continuous("", limits=c(0,3.5)) +
  theme(plot.title = element_text(size = 14, face = "bold"),
        legend.title=element_text(size=12,  face = "bold"), 
        legend.text=element_text(size=10)) +
  annotate("segment", x=-Inf, xend=Inf, y=-Inf, yend=-Inf)+
  annotate("segment", x=-Inf, xend=-Inf, y=-Inf, yend=Inf) +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank())
```

Specifying the title and legends:
```{r} 
pt_d <- pt_c + guides(fill = guide_legend("Significant", reverse = TRUE, nrow = 3)) +
  theme(legend.position="right") + labs(title="Public transport")
```

Add the number of intervention components (**n_interventions**) above each bar:
```{r} 
pt_final <- pt_d +  geom_text(aes(label=n_interventions), position=position_dodge(width=0.9), vjust=-0.4)
```

The final plot should look like this:
```{r, fig.width=10,fig.height=6, warning=FALSE}
pt_final
```
<br><br>

***

# Creating the final combined harvest plot
We then use the ggarrange function to plot both the Driving and Public transport harvest plot figures together:
```{r, fig.width=10,fig.height=12}
figure <- ggarrange(drive_final, pt_final, hjust = -0.5,
                    ncol = 1, nrow = 2)
figure
```
